Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2020-04-10 pkg/circtree/inst/wind_paper/draft/JRSSC/rebuttal.tex

\documentclass[english, noconfig]{uibklttr}

\newcommand{\section}[1]{{\Large{\textbf{#1}}}}
\newcommand{\subsection}[1]{{\large \textbf{#1}}}
\newcommand{\subsubsection}[1]{{\large \textbf{#1}}}
\newcommand{\tightlist}[0]{}

% Some colors
\definecolor{myblue}{rgb}{0.2,0.2,0.9}
\definecolor{mygray}{rgb}{0.7,0.7,0.7}

\makeatletter
\newenvironment{thebibliography}[1]
     {\list{\@biblabel{\@arabic\c@enumiv}}%
           {\settowidth\labelwidth{\@biblabel{#1}}%
            \leftmargin\labelwidth
            \advance\leftmargin\labelsep
            \usecounter{enumiv}%
            \let\p@enumiv\@empty
            \renewcommand\theenumiv{\@arabic\c@enumiv}}%
      \sloppy
      \clubpenalty4000
      \@clubpenalty \clubpenalty
      \widowpenalty4000%
      \sfcode`\.\@m}
     {\def\@noitemerr
       {\@latex@warning{Empty `thebibliography' environment}}%
      \endlist}
\newcommand\newblock{\hskip .11em\@plus.33em\@minus.07em}
\makeatother
\usepackage{natbib}

%\usepackage{subcaption}
%\captionsetup{font={bf,small},skip=0.25\baselineskip}
%\captionsetup[subfigure]{font={bf,small}, skip=1pt, singlelinecheck=false}

\usepackage{amsmath}

\usepackage{float} % For figure placements
\usepackage{enumitem}
\setitemize{itemsep=0em}
\setlist{itemsep = 0em}

\usepackage{parskip}
\setlength{\parskip}{.4em}

% fake-float with caption for figures
\usepackage{caption}
\usepackage{newfloat}
\DeclareFloatingEnvironment[fileext=lot]{figure}
\DeclareCaptionFont{myblue}{\color{myblue}}
\captionsetup{labelfont={bf},textfont={it}}

% short re
\newcommand{\sre}[1]{
    \begin{color}{uibkgraym}#1\end{color}%
}
% long re
\newenvironment{re}{
    \begin{color}{uibkgraym}
        \itshape
}{
    \end{color}
}
\newcommand{\todo}[1]{
    \begin{color}{red}\textbf{TODO:} #1\end{color}%
}
\newcommand{\note}[1]{
    \begin{color}{orange}\textbf{Reto's comment:} #1\end{color}%
}
\newcommand{\see}[2]{Page \textbf{#1}, line \textbf{#2}.}
\newcommand{\msee}[3]{Page \textbf{#1}, lines \textbf{#2--#3}.}
\newcommand{\fig}[1]{\textit{now Fig~#1}}
\newcommand{\nsec}[1]{\textit{now Sec~#1}}

\begin{document}

\setkomavar{subject}{Revision of `npg-2019-49'}
\setkomavar{signature}{Moritz Lang\\Corresponding Author}

\setkomavar{fromname}[Name]{Moritz Lang}
\setkomavar{fromemail}{Moritz.Lang@uibk.ac.at}

\setkomavar{fromaddress}{University of Innsbruck, Department of Statistics, Universit\"atsstrasse
15, 6020 Innsbruck, Austria}

%\begin{letter}[fromemail]{To Maxime Taillardat}
%\opening{Dear Christopher Paciorek}
%
%We thank you, as well as both reviewers, for the detailed and fruitful feedback
%regarding our manuscript ``\textbf{Bivariate Gaussian models for wind vectors in a
%distributional regression framework}''.
%
%\vspace{5mm}
%
%We have carefully revised our manuscript according to your comments and
%suggestions. The most substantial changes are the following:
%
%\begin{itemize}
%\item
%An additional comparison to the work of \citet{schuhen.etal:2012} and of
%\citet{pinson:2012} has been included. Both comparisons nicely confirm our
%previous results. To keep the manuscript as clear as possible, we describe the
%comparisons in the appendix and provide short summaries of the main finding
%in the discussion section.
%
%\item
%The introduction of the logarithmic score has been
%extended. We now explicitly derive the employed formulation of the logarithmic
%score and provide additional references.
%
%\item
%We have added more information on the estimation procedure and describe the key
%steps involved in the model fitting. Please also remark that for consistency we
%have changed `mean and variance` to `location and scale parameter` within the
%abstract and the introductory part of the discussion.
%\end{itemize}
%
%\vspace{5mm}
%
%On the following pages a point-to-point response to both reviewers and to the
%editor will be given. The attached manuscript highlights the changes in the
%text in blue color.
%
%\closing{Your sincerely,}
%
%\end{letter}
%
%\newpage

\section{Editor: Prof. Peter Smith - Comments to the Author}

\begin{itemize}

\item Please can you rewrite the Summary to give the application more prominence.
Currently, the last sentence reads to me like an afterthought. Furthermore, I
would prefer you to use the term ‘application’ rather than ‘case study’
throughout the paper, since in Applied Statistics papers the methodological
developments should be motivated by the application.

\item Page 2, line 46: ‘… structure can capture non-additive …’?

\item Page 2, lines 54 and 55: ‘… mean, whereas modelling the conditional … distribution would also allow uncertainty in the forecast to be estimated.’?

\item Page 6, line 16 and Equation (2.6): bold the vector $\beta$.

\item Page 6, lines 29 and 30: ‘In general, for additive models a proper model …’?

\item Equations (3.1) to (3.4) and elsewhere: I find the use of multiple letters for a single superscript or subscript awkward and prefer to keep lower case letters for indices, so please consider replacing them, for example, T for tree and F for forest. Also for Equation (4.2), how about using parentheses rather than subscripts?

\item Equation (3.4): bold vector $z$.

\item Page 9, line 39: delete ‘respectively’.

\item Page 15, line 23: replace ‘leveraged’ with ‘used’?

\end{itemize}

\newpage

\section{Associate Editor - Comments to the Author}

Both referees agree this is a good paper, and have only a few suggestions for
changes, so I recommend minor revisions to address these.

I have two short suggestions in addition to this:

\begin{itemize}
\item Can you comment on the typical computational requirements of the various
methods used in your application, in terms of hardware required and length of
computing time.

\item Is it possible to include a link to computer code to reproduce your
analysis? This is not essential for publication but would be helpful for
readers who wish to perform similar analyses. The R package you have provided
for your method is very helpful in this regard.  
\end{itemize}

\newpage

\section{Referee: 1 - Comments to the Author}

I found Circular Regression Trees and Forests to be a very nicely written
paper, with a meaningful contribution to the literature on regression models
for a circular response variable.

I have very few suggestions to the authors; but here are two:

\begin{itemize}
\item On page 2, what is the reference for the phrase "Additionally, they
developed a combined ... version ...".

\item On page 11, regarding "As the implementation cannot handle circular
covariates ...": It could handle circular covariates in a fashion similar to
Lund (1999), using a rotational or reflective relationship, or modeling the
association with a trigonometric polynomial. Adding this to their comparison
model, or at least mentioning the possibility of doing so, would be more
appropriate than stating that circular covariates cannot be accommodated.
\end{itemize}

\vspace{0.5cm}
References

Lund UJ (1999). ``Least Circular Distance Regression for Directional Data.''
\emph{Journal of Applied Statistics}, \textbf{26}\,(6), 723--733.
doi:10.1080/02664769922160.

\newpage

\section{Referee: 2 - Comments to the Author}

This is a well-written paper on using circular regression trees for analyzing a
specific type of circular data. The probabilistic approach is well-founded. The
experiments on wind direction data are well organized and the results are
thoroughly analyzed.

\begin{itemize}
\item The authors could address model order determination problem using the
Bayes Information Criterion (Schwartz, 1978, Annals of Statistics).

\item I would like to point out related works on circular regression models for
shape representation (Kashyap and Chellappa) and circular random forests for
vehicle direction estimation (Hara and Chellappa) given below. I feel there is
some overlap between this paper and the work of Hara and Chellappa on using
circular statistics for vehicle orientation estimation. The authors may include
some comparisons with Hara, Chellappa work.
\end{itemize}

\vspace{0.5cm}
References

Hara K, Chellappa R (2017). ``Growing Regression Tree Forests by
Classification for Continuous Object Pose Estimation.'' \emph{International
Journal of Computer Vision}, \textbf{122}\,(2), 292--312.
doi:10.1007/s11263-016-0942-1.

Kashyap R, Chellappa R (1981). ``Stochastic Models for Closed Boundary
Analysis: Representation and Reconstruction.'' \emph{IEEE Transactions on
Information Theory}, \textbf{27}\,(5), 627--637. doi:10.1109/TIT.1981.1056390.

\end{document}

### 2020-04-10 pkg/circtree/inst/wind_paper/draft/JRSSC/draft_jrssc_submitted_2020_01_09.tex
Create diff option in makefile
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3006 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7afc18d26503bf349a371078d37fa6d921a0e74c

\documentclass[nojss,shortnames]{jss}

%% packages
\usepackage{amstext,amsfonts,amsmath,bm,thumbpdf,lmodern,hyperref}
\usepackage[all]{hypcap}

@@

%% need no \usepackage{Sweave}
\SweaveOpts{engine = R, concordance = FALSE, eps = FALSE, keep.source = TRUE, eval = TRUE}

%% commands
\newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
\renewcommand{\Prob}{\mathbb{P} }
\renewcommand{\E}{\mathbb{E}}
\newcommand{\V}{\mathbb{V}}
\newcommand{\Var}{\mathbb{V}}
\newcommand{\R}{\mathbb{R} }
\newcommand{\N}{\mathbb{N} }
\newcommand{\C}{\mathbb{C} }
\newcommand{\argmin}{\operatorname{argmin}\displaylimits}
\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
\newcommand{\LS}{\mathcal{L}_n}
\newcommand{\TS}{\mathcal{T}_n}
\newcommand{\LSc}{\mathcal{L}_{\text{comb},n}}
\newcommand{\LSbc}{\mathcal{L}^*_{\text{comb},n}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\yn}{y_{\text{new}}}
\newcommand{\z}{\mathbf{z}}
\newcommand{\X}{\mathbf{X}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\sX}{\mathcal{X}}
\newcommand{\sY}{\mathcal{Y}}
\newcommand{\T}{\mathbf{T}}
\newcommand{\x}{\mathbf{x}}
\renewcommand{\a}{\mathbf{a}}
\newcommand{\xn}{\mathbf{x}_{\text{new}}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\ws}{\mathbf{w}_\cdot}
\renewcommand{\t}{\mathbf{t}}
\newcommand{\M}{\mathbf{M}}
\renewcommand{\vec}{\text{vec}}
\newcommand{\B}{\mathbf{B}}
\newcommand{\K}{\mathbf{K}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\I}{\mathbf{I}}
\newcommand{\bS}{\mathbf{S}}
\newcommand{\cellx}{\pi_n[\x]}
\newcommand{\partn}{\pi_n(\mathcal{L}_n)}
\newcommand{\err}{\text{Err}}
\newcommand{\ea}{\widehat{\text{Err}}^{(a)}}
\newcommand{\ecv}{\widehat{\text{Err}}^{(cv1)}}
\newcommand{\ecvten}{\widehat{\text{Err}}^{(cv10)}}
\newcommand{\eone}{\widehat{\text{Err}}^{(1)}}
\newcommand{\eplus}{\widehat{\text{Err}}^{(.632+)}}
\newcommand{\eoob}{\widehat{\text{Err}}^{(oob)}}
\newcommand{\bft}{\mathbf{t}}

\definecolor{changes}{rgb}{1,1,0.7}
\newcommand{\hlc}[1]{{\sethlcolor{changes}\hl{#1}}}

\numberwithin{equation}{section}

<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
library("disttree")
library("circtree")

## HCL palette
pal <- hcl(c(10, 128, 260, 290, 50), 100, 50)
pallight <- hcl(c(10, 128, 260, 290, 70), 100, 50, alpha = 0.25)
transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)

# set function for parallelization
applyfun <- function(X, FUN, ...) parallel::mclapply(X, FUN, ..., mc.cores = pmax(1, parallel::detectCores() - 1))

@

\title{Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting}
\Shorttitle{Circular Regression Trees and Forests}

\author{Moritz N. Lang\\Universit\"at Innsbruck
   \And Lisa Schlosser\\Universit\"at Innsbruck
   \And Torsten Hothorn\\Universit\"at Z\"urich
   \AND Georg~J.~Mayr\\Universit\"at Innsbruck
   \And Reto Stauffer\\Universit\"at Innsbruck
   \And Achim Zeileis\\Universit\"at Innsbruck}
\Plainauthor{Lang, Schlosser, Hothorn, Mayr, Stauffer, Zeileis}

\Abstract{
While circular data occur in a wide range of scientific fields, the methodology
for distributional modeling and probabilistic forecasting of circular response
variables is rather limited. Most of the existing methods are built on the
framework of generalized linear and additive models, which are often
challenging to optimize and interpret. Therefore, building on previous ideas
for trees modeling circular means, we suggest a distributional approach for
regression trees and random forests yielding probabilistic forecasts based on
the von~Mises distribution. The resulting tree-based models simplify the
estimation process by using the available covariates for partitioning the data
into sufficiently homogeneous subgroups so that a simple von~Mises distribution
without further covariates can be fitted to the circular response in each
subgroup. These circular regression trees are straightforward to interpret, can
capture nonlinear effects and interactions, and automatically select the
relevant covariates that are associated with either location and/or scale
changes in the von~Mises distribution. Combining an ensemble of circular
regression trees to a circular regression forest can regularize and smooth the
covariate effects. The new methods are evaluated in a case study on
probabilistic wind direction forecasting at two Austrian airports, considering
other common approaches as a benchmark.
}

\Keywords{circular data, regression trees, random forests, distributional regression, von~Mises distribution, probabilistic forecasting}

\Address{
  Moritz N. Lang, Lisa Schlosser, Achim Zeileis \\
  Universit\"at Innsbruck \\
  Department of Statistics \\
  Faculty of Economics and Statistics \\
  Universit\"atsstr.~15 \\
  6020 Innsbruck, Austria \\
  E-mail: \email{Moritz.Lang@uibk.ac.at}, \\
  \phantom{E-mail: }\email{Lisa.Schlosser@uibk.ac.at},\\
  \phantom{E-mail: }\email{Achim.Zeileis@R-project.org} \\
  URL: \url{https://www.uibk.ac.at/statistics/personal/moritz-lang/},\\
  \phantom{URL: }\url{https://www.uibk.ac.at/statistics/personal/schlosser-lisa/},\\
  \phantom{URL: }\url{https://eeecon.uibk.ac.at/~zeileis/}\\

  Reto Stauffer \\
  Universit\"at Innsbruck \\
  Digital Science Center and \\ 
  Department of Statistics \\
  Faculty of Economics and Statistics \\
  Universit\"atsstr.~15 \\
  6020 Innsbruck, Austria \\
  E-mail: \email{Reto.Stauffer@uibk.ac.at} \\
  URL: \url{https://retostauffer.org/}\\

  Georg~J.~Mayr\\
  Universit\"at Innsbruck \\
  Department of Atmospheric and Cryospheric Science \\
  Faculty of Geo- and Atmospheric Sciences \\
  Innrain~52f \\
  6020 Innsbruck, Austria \\
  E-mail: \email{Georg.Mayr@uibk.ac.at} \\
  URL: \url{https://www.uibk.ac.at/acinn/people/georg-mayr}\\

  Torsten Hothorn\\
  Universit\"at Z\"urich \\
  Institut f\"ur Epidemiologie, Biostatistik und Pr\"avention \\
  Hirschengraben 84\\
  CH-8001 Z\"urich, Switzerland \\
  E-mail: \email{Torsten.Hothorn@R-project.org}\\
  URL: \url{http://user.math.uzh.ch/hothorn/}\\

}

@@

@@

\newpage

@@

% -------------------------------------------------------------------
% Probabilistic circular modeling
% -------------------------------------------------------------------
\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a probability
distribution which accounts for the periodicity of circular data. Generally,
this feature can be obtained by ``wrapping'' the probability density function
of any continuous distribution around the unit circle \citep{Mardia+Jupp:1999}.
In that way, the wrapped Cauchy distribution or the wrapped normal distribution
can be employed to model symmetric unimodal circular data. A close
approximation to the wrapped normal distribution that is mathematically simpler
and hence easier to use is provided by the von~Mises distribution
\citep{Fisher:1993}, a purely circular distribution which is also known as
``the circular normal distribution'' and is a common choice for probabilistic
modeling of circular data. Based on a location parameter $\mu \in [0, 2\pi)$
and a concentration parameter $\kappa > 0$ the density of the von~Mises
distribution for an observation~\mbox{$y \in [0, 2\pi)$} is given by
%%
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)},\label{schlosser:equ_vm}
\end{equation}
%%
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
$0$ (see, e.g., \citealt{Mardia+Zemroch:1975b}, \citealt{Jammalamadaka+Sengupta:2001}, or
\citealt{Ley+Verdebout:2017} for a more detailed overview).

@@

The corresponding log-likelihood function is defined as
%%
\begin{equation}
    \ell(\mu, \kappa; y) = \log(f_\mathrm{vM}(y;\mu, \kappa))
    = -\log(2 \pi I_0(\kappa)) + \kappa \cos(y - \mu).\label{equ:vM:loglik}
\end{equation}
%%
To fit a probabilistic model $\text{vM}(y; \mu, \kappa)$ to a circular response
$y$, the distribution parameters~$\mu$ and~$\kappa$ need to be estimated. This
can be done by maximizing the log-likelihood function~$\ell(\mu, \kappa; y)$
%%
\begin{equation}\label{eq:optim}
  (\hat{\mu},\hat{\kappa}) = \argmax_{\mu,\kappa} \sum_{i=1}^n \ell(\mu, \kappa; y_i) 
\end{equation}
%%
yielding maximum likelihood estimators $\hat{\mu}$ and $\hat{\kappa}$ such that
a fully specified distributional model is fitted to the learning data
$\{y_i\}_{i=1,\ldots,n}$.

@@

\subsection{Circular regression forests}
\label{sec:circforest} 
A natural extension of (circular) regression trees are ensembles or forests
that can improve forecasts by regularizing and stabilizing the model. Random
forests introduced by \cite{Breiman:2001} average the predictions of an
ensemble of trees, each built on a subsample or bootstrap of the original data.
A generalization of this strategy is to obtain weighted predictions by adaptive
local likelihood estimation of the distributional parameters
\citep[Section~2.3. of][]{Schlosser+Hothorn+Stauffer:2019,
Hothorn+Zeileis:2017}. More specifically, for each (possibly new)
observation~$\bm{z}$ a set of averaged ``nearest neighbor'' weights
$w_i^\text{forest}(\bm{z})$ is obtained that is based on the number of trees in
which $\bm{z}$ is assigned to the same terminal node as each learning
observation~\mbox{$y_i, i \in \{1,\ldots,n\}$}. Hence, for a forest of $T$ trees, the
weights are calculated via
%%
\begin{equation}
  w^{\text{forest}}_i(\bm{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
  \frac{\mathbf{1}((\bm{z}_i \in \mathcal{B}^t_b) \land (\bm{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
\end{equation}
%%
where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
segment of the $t$-th tree. Therefore, similar observations ending up more
often in the same terminal node have higher weights and thus a stronger
influence in the weighted maximum likelihood estimation.

@@

Therefore, the resulting parameter estimates can smoothly adapt to the given
covariates $\bm{z}$ whereas $w_i(\bm{z}) = 1$ would correspond to the
unweighted full-sample estimates and \mbox{$w_i(\bm{z}) \in \{0, 1\}$} corresponds to
the subgroup selection from a tree. Thus, circular regression forests can
capture both smooth and abrupt changes, while covariates and possible
interactions are selected automatically and do not explicitly need to be
specified beforehand.

@@

%%
\begin{table}[t!]
 \caption[Table caption text]{Overview of the data sets employed in the case
    study: For Innsbruck and Vienna, various meteorological variables and derived
    quantities of these are considered at the respective stations, located either
    directly at the airports or in their vicinities.}
  \label{tab:data}
  \begin{center}
    \begin{tabular}{l  l}
      \hline
      Data components               & Description \\
      \hline
      Temporal information:         & Time of the day, day of the year \\
      \noalign{\vskip 1mm}   
      Meteorological variables:     & Wind direction, wind (gust) speed, \\
                                    & (reduced) air pressure, relative humidity, \\
                                    & temperature\\ 
      \noalign{\vskip 1mm}                           
      Derived quantities:           & 3-hourly means/minima/maxima, \\
                                    & 1-hourly and 3-hourly temporal changes, \\
                                    & spatial differences towards the airport\\
      \noalign{\vskip 1mm}                              
      Weather stations (Innsbruck): & 4 stations at the airport, as well as\\
                                    & Igls, Kematen, Kufstein, Landeck, Patscherkofel, \\
                                    & and Steinach\\
      \noalign{\vskip 1mm}
      Weather stations (Vienna):    & 9 stations at the airport, as well as\\
                                    & Arsenal, Donaufeld, Exelberg, G\"anserndorf, \\
                                    & Gro{\ss}-Enzersdorf, Gumpoldskirchen, Hohe Warte,\\
                                    & Innere Stadt, Jubil\"aumswarte, Mariabrunn, \\
                                    & Seibersdorf, Unterlaa, and Wolkersdorf\\
      \hline
    \end{tabular}
  \end{center}
\end{table}
%%

@@

    tree is built with the newly developed \textsf{R} package \pkg{circtree} employing the

@@

    2000 observations in each terminal node (argument \code{minbucket}).

@@

    \pkg{circtree}. Each of these trees is again built by the CTree algorithm on a

@@

    be applied in \pkg{circtree} by setting the argument \code{mtry} to the total

@@

    classes (argument \code{nmax = c(yx = Inf, z = 50)}). Contrary to a

>>>>

\documentclass{statsoc}

\usepackage[a4paper]{geometry}
%\usepackage{graphicx}

%%% packages
\usepackage{amstext,amsfonts,amsmath,bm,thumbpdf,lmodern,hyperref,natbib}
%\usepackage[all]{hypcap} %% FAILS IN JRSSC

@@

%% need no \usepackage{Sweave}

%% commands
%\newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
%\newcommand{\Prob}{\mathbb{P} }
%\newcommand{\E}{\mathbb{E}}
%\newcommand{\V}{\mathbb{V}}
%\newcommand{\Var}{\mathbb{V}}
%\newcommand{\R}{\mathbb{R} }
%\newcommand{\N}{\mathbb{N} }
%\newcommand{\C}{\mathbb{C} }
\newcommand{\argmin}{\operatorname{argmin}\displaylimits}
\newcommand{\argmax}{\operatorname{argmax}\displaylimits}
%\newcommand{\LS}{\mathcal{L}_n}
%\newcommand{\TS}{\mathcal{T}_n}
%\newcommand{\LSc}{\mathcal{L}_{\text{comb},n}}
%\newcommand{\LSbc}{\mathcal{L}^*_{\text{comb},n}}
%\newcommand{\F}{\mathcal{F}}
%\newcommand{\A}{\mathcal{A}}
%\newcommand{\yn}{y_{\text{new}}}
%\newcommand{\z}{\mathbf{z}}
%\newcommand{\X}{\mathbf{X}}
%\newcommand{\Y}{\mathbf{Y}}
%\newcommand{\sX}{\mathcal{X}}
%\newcommand{\sY}{\mathcal{Y}}
%\newcommand{\T}{\mathbf{T}}
%\newcommand{\x}{\mathbf{x}}
%\renewcommand{\a}{\mathbf{a}}
%\newcommand{\xn}{\mathbf{x}_{\text{new}}}
%\newcommand{\y}{\mathbf{y}}
%\newcommand{\w}{\mathbf{w}}
%\newcommand{\ws}{\mathbf{w}_\cdot}
%\renewcommand{\t}{\mathbf{t}}
%\newcommand{\M}{\mathbf{M}}
%\renewcommand{\vec}{\text{vec}}
%\newcommand{\B}{\mathbf{B}}
%\newcommand{\K}{\mathbf{K}}
%\newcommand{\W}{\mathbf{W}}
%\newcommand{\D}{\mathbf{D}}
\newcommand{\I}{\mathbf{I}}
%\newcommand{\bS}{\mathbf{S}}
%\newcommand{\cellx}{\pi_n[\x]}
%\newcommand{\partn}{\pi_n(\mathcal{L}_n)}
%\newcommand{\err}{\text{Err}}
%\newcommand{\ea}{\widehat{\text{Err}}^{(a)}}
%\newcommand{\ecv}{\widehat{\text{Err}}^{(cv1)}}
%\newcommand{\ecvten}{\widehat{\text{Err}}^{(cv10)}}
%\newcommand{\eone}{\widehat{\text{Err}}^{(1)}}
%\newcommand{\eplus}{\widehat{\text{Err}}^{(.632+)}}
%\newcommand{\eoob}{\widehat{\text{Err}}^{(oob)}}
%\newcommand{\bft}{\mathbf{t}}

%\definecolor{changes}{rgb}{1,1,0.7}
%\newcommand{\hlc}[1]{{\sethlcolor{changes}\hl{#1}}}
%
%\numberwithin{equation}{section}

\graphicspath{{../}}

\title[Circular Regression Trees and Forests]{Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting}

\author{Moritz N. Lang}
\address{
  Universit\"at Innsbruck,
  Innsbruck, Austria.}
\email{Moritz.Lang@uibk.ac.at}
\author{Lisa Schlosser}
\address{
  Universit\"at Innsbruck,
  Innsbruck, Austria}
\author{Torsten Hothorn.}
\address{
  Universit\"at Z\"urich,
  Z\"urich, Switzerland.}
\author{Georg~J.~Mayr}
\address{
  Universit\"at Innsbruck,
  Innsbruck, Austria.}
\author{Reto Stauffer}
\address{
  Universit\"at Innsbruck,
  Innsbruck, Austria.}
\author[Lang, Schlosser, Hothorn, Mayr, Stauffer, Zeileis]{Achim Zeileis}
\address{
  Universit\"at Innsbruck,
  Innsbruck, Austria.}

@@

\begin{abstract}
While circular data occur in a wide range of scientific fields, the methodology
for distributional modeling and probabilistic forecasting of circular response
variables is rather limited. Most of the existing methods are built on the
framework of generalized linear and additive models, which are often
challenging to optimize and interpret. Therefore, building on previous ideas
for trees modeling circular means, we suggest a distributional approach for
regression trees and random forests yielding probabilistic forecasts based on
the von~Mises distribution. The resulting tree-based models simplify the
estimation process by using the available covariates for partitioning the data
into sufficiently homogeneous subgroups so that a simple von~Mises distribution
without further covariates can be fitted to the circular response in each
subgroup. These circular regression trees are straightforward to interpret, can
capture nonlinear effects and interactions, and automatically select the
relevant covariates that are associated with either location and/or scale
changes in the von~Mises distribution. Combining an ensemble of circular
regression trees to a circular regression forest can regularize and smooth the
covariate effects. The new methods are evaluated in a case study on
probabilistic wind direction forecasting at two Austrian airports, considering
other common approaches as a benchmark.
\end{abstract}

\keywords{Circular data; Regression trees; Random forests; Distributional regression; von~Mises distribution}

@@

@@

% -------------------------------------------------------------------
% Probabilistic circular modeling
% -------------------------------------------------------------------
\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a probability
distribution which accounts for the periodicity of circular data. Generally,
this feature can be obtained by ``wrapping'' the probability density function
of any continuous distribution around the unit circle \citep{Mardia+Jupp:1999}.
In that way, the wrapped Cauchy distribution or the wrapped normal distribution
can be employed to model symmetric unimodal circular data. A close
approximation to the wrapped normal distribution that is mathematically simpler
and hence easier to use is provided by the von~Mises distribution
\citep{Fisher:1993}, a purely circular distribution which is also known as
``the circular normal distribution'' and is a common choice for probabilistic
modeling of circular data. Based on a location parameter $\mu \in [0, 2\pi)$
and a concentration parameter $\kappa > 0$ the density of the von~Mises
distribution for an observation $y \in [0, 2\pi)$ is given by
%%
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)},\label{schlosser:equ_vm}
\end{equation}
%%
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
$0$ (see, e.g., \citealt{Mardia+Zemroch:1975b}, \citealt{Jammalamadaka+Sengupta:2001}, or
\citealt{Ley+Verdebout:2017} for a more detailed overview).

@@

The corresponding log-likelihood function is defined as
%%
\begin{equation}
    \ell(\mu, \kappa; y) = \log(f_\mathrm{vM}(y;\mu, \kappa))
    = -\log(2 \pi I_0(\kappa)) + \kappa \cos(y - \mu).\label{equ:vM:loglik}
\end{equation}
%%
To fit a probabilistic model $\text{vM}(y; \mu, \kappa)$ to a circular response
$y$, the distribution parameters $\mu$ and $\kappa$ need to be estimated. This
can be done by maximizing the log-likelihood function $\ell(\mu, \kappa; y)$
%%
\begin{equation}\label{eq:optim}
  (\hat{\mu},\hat{\kappa}) = \argmax_{\mu,\kappa} \sum_{i=1}^n \ell(\mu, \kappa; y_i) 
\end{equation}
%%
yielding maximum likelihood estimators $\hat{\mu}$ and $\hat{\kappa}$ such that
a fully specified distributional model is fitted to the learning data
$\{y_i\}_{i=1,\ldots,n}$.

@@

\subsection{Circular regression forests}
\label{sec:circforest} 
A natural extension of (circular) regression trees are ensembles or forests
that can improve forecasts by regularizing and stabilizing the model. Random
forests introduced by \cite{Breiman:2001} average the predictions of an
ensemble of trees, each built on a subsample or bootstrap of the original data.
A generalization of this strategy is to obtain weighted predictions by adaptive
local likelihood estimation of the distributional parameters
\citep[Section~2.3. of][]{Schlosser+Hothorn+Stauffer:2019,
Hothorn+Zeileis:2017}. More specifically, for each (possibly new)
observation~$\bm{z}$ a set of averaged ``nearest neighbor'' weights
$w_i^\text{forest}(\bm{z})$ is obtained that is based on the number of trees in
which $\bm{z}$ is assigned to the same terminal node as each learning
observation $y_i, i \in \{1,\ldots,n\}$. Hence, for a forest of $T$ trees, the
weights are calculated via
%%
\begin{equation}
  w^{\text{forest}}_i(\bm{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
  \frac{\mathbf{1}((\bm{z}_i \in \mathcal{B}^t_b) \land (\bm{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
\end{equation}
%%
where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
segment of the $t$-th tree. Therefore, similar observations ending up more
often in the same terminal node have higher weights and thus a stronger
influence in the weighted maximum likelihood estimation.

@@

Therefore, the resulting parameter estimates can smoothly adapt to the given
covariates $\bm{z}$ whereas $w_i(\bm{z}) = 1$ would correspond to the
unweighted full-sample estimates and $w_i(\bm{z}) \in \{0, 1\}$ corresponds to
the subgroup selection from a tree. Thus, circular regression forests can
capture both smooth and abrupt changes, while covariates and possible
interactions are selected automatically and do not explicitly need to be
specified beforehand.

@@

%%
\begin{table}[t!]
 \caption[Table caption text]{Overview of the data sets employed in the case
    study: For Innsbruck and Vienna, various meteorological variables and derived
    quantities of these are considered at the respective stations, located either
    directly at the airports or in their vicinities.}
  \label{tab:data}
  \centering
    \begin{tabular}{l  l}
      \hline
      Data components               & Description \\
      \hline
      Temporal information:         & Time of the day, day of the year \\
      \noalign{\vskip 1mm}   
      Meteorological variables:     & Wind direction, wind (gust) speed, \\
                                    & (reduced) air pressure, relative humidity, \\
                                    & temperature\\ 
      \noalign{\vskip 1mm}                           
      Derived quantities:           & 3-hourly means/minima/maxima, \\
                                    & 1-hourly and 3-hourly temporal changes, \\
                                    & spatial differences towards the airport\\
      \noalign{\vskip 1mm}                              
      Weather stations (Innsbruck): & 4 stations at the airport, as well as\\
                                    & Igls, Kematen, Kufstein, Landeck, Patscherkofel, \\
                                    & and Steinach\\
      \noalign{\vskip 1mm}
      Weather stations (Vienna):    & 9 stations at the airport, as well as\\
                                    & Arsenal, Donaufeld, Exelberg, G\"anserndorf, \\
                                    & Gro{\ss}-Enzersdorf, Gumpoldskirchen, Hohe Warte,\\
                                    & Innere Stadt, Jubil\"aumswarte, Mariabrunn, \\
                                    & Seibersdorf, Unterlaa, and Wolkersdorf\\
      \hline
    \end{tabular}
\end{table}
%%

@@

    tree is built with the newly developed \textsf{R} package \textbf{circtree} employing the

@@

    2000 observations in each terminal node (argument \texttt{minbucket}).

@@

    \textbf{circtree}. Each of these trees is again built by the CTree algorithm on a

@@

    be applied in \textbf{circtree} by setting the argument \texttt{mtry} to the total

@@

    classes (argument \texttt{nmax = c(yx = Inf, z = 50)}). Contrary to a
